---
title: "WGCNA GO Semantic Similarity"
author: "Expression_Quant"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    keep_md: yes
---



```{r setup, include=FALSE}
# The following setting is important, do not omit.
knitr::opts_chunk$set(echo = TRUE,
                      fig.height = 8,
                      fig.width = 12,
                      dev = c("png", "pdf"),
                      dpi = 1000)
options(scipen = 999) # Prevent scientific notation
rm(list = ls()) # Clear environment
```

*Note: Open scripts.Rproj first, then script. To easily use relative paths, click the down button next to knit and then click "Knit Directory --> Project Directory". This should make loading and saving files much easier.*



# GO Semantic Similarity

Perform GO Semantic Similarity analysis to gain additional insights into our WGCNA enriched GO terms, using results from WGCNA analysis (i.e., after running the `2a-WGCNA.Rmd` script).
<span style="color:red">Red text</span> indicates regions that require the user to modify.
Regardless, the user should check over all code blocks to ensure that everything is running correctly.



## 1. Load packages

Load packages
```{r load_packages, message=FALSE, warning=FALSE}
library(tidyverse)
library(org.Hs.eg.db)
library(simplifyEnrichment)
library(magick)
```



## 2. Set variables

<span style="color:red">**Change file names and conditions where appropriate.**</span>
```{r variables}
# Input unfiltered data
GO_term.file <- "salmon.numreads.WGCNA_results.GOsig.tsv"
```



## 3. Load, clean, and pre-processing datasets

Load GO data and select only the GO_IDs associated with BP, MF, and CC terms.
```{r load_GO_terms}
GO <- read.csv(GO_term.file, header=TRUE, sep="\t")

# Print stats
print(paste("Number sig BP terms: ", nrow(filter(GO, ontology=="BP")), sep=''))
print(paste("Number sig MF terms: ", nrow(filter(GO, ontology=="MF")), sep=''))
print(paste("Number sig CC terms: ", nrow(filter(GO, ontology=="CC")), sep=''))

# Extract BP, MF, and CC separatly
GO.BP <- GO %>% filter(ontology == "BP")
BP <- GO.BP$category
GO.MF <- GO %>% filter(ontology == "MF")
MF <- GO.MF$category
GO.CC <- GO %>% filter(ontology == "CC")
CC <- GO.CC$category
```



## 4. Calculate a similarity matrix and save the output

```{r BP_simplifyGO}
simplifyGO(GO_similarity(BP, ont = "BP", db = "org.Hs.eg.db"),
           word_cloud_grob_param = list(max_width = 50),
           max_words=20
          )
```


```{r MF_simplifyGO}
simplifyGO(GO_similarity(MF, ont = "MF", db = "org.Hs.eg.db"),
           word_cloud_grob_param = list(max_width = 50),
           max_words=20
          )
```


```{r CC_simplifyGO}
simplifyGO(GO_similarity(CC, ont = "CC", db = "org.Hs.eg.db"),
           word_cloud_grob_param = list(max_width = 50),
           max_words=20
          )
```



## 5. Plot term significance values

```{r plot_BP_significance}
GO.BP %>%
  mutate(term = fct_reorder(term, over_represented_pvalue, .desc = TRUE)) %>%
  ggplot(aes(x=term, y=over_represented_pvalue) ) +
      geom_segment( aes(x=term ,xend=term, y=0, yend=over_represented_pvalue), color="grey") +
      geom_point(size=3, color="#69b3a2") +
      coord_flip() +
      theme(
        panel.grid.minor.y = element_blank(),
        panel.grid.major.y = element_blank(),
        legend.position="none"
      ) +
  xlab("") +
  ylab("over_represented_pvalue") +
  ggtitle("Adult") + #add a main title
  theme(plot.title = element_text(face = 'bold',
                                  size = 12,
                                  hjust = 0)) +
  theme_bw() + #Set background color
  theme(panel.border = element_blank(), # Set border
                     panel.grid.major = element_blank(), #Set major gridlines
                     panel.grid.minor = element_blank(), #Set minor gridlines
                     axis.line = element_line(colour = "black"), #Set axes color
        plot.background=element_blank(),#Set the plot background
        legend.position="none")
```


```{r plot_MF_significance}
GO.MF %>%
  mutate(term = fct_reorder(term, over_represented_pvalue, .desc = TRUE)) %>%
  ggplot(aes(x=term, y=over_represented_pvalue) ) +
      geom_segment( aes(x=term ,xend=term, y=0, yend=over_represented_pvalue), color="grey") +
      geom_point(size=3, color="#69b3a2") +
      coord_flip() +
      theme(
        panel.grid.minor.y = element_blank(),
        panel.grid.major.y = element_blank(),
        legend.position="none"
      ) +
  xlab("") +
  ylab("over_represented_pvalue") +
  ggtitle("Adult") + #add a main title
  theme(plot.title = element_text(face = 'bold',
                                  size = 12,
                                  hjust = 0)) +
  theme_bw() + #Set background color
  theme(panel.border = element_blank(), # Set border
                     panel.grid.major = element_blank(), #Set major gridlines
                     panel.grid.minor = element_blank(), #Set minor gridlines
                     axis.line = element_line(colour = "black"), #Set axes color
        plot.background=element_blank(),#Set the plot background
        legend.position="none")
```


```{r plot_CC_significance}
GO.CC %>%
  mutate(term = fct_reorder(term, over_represented_pvalue, .desc = TRUE)) %>%
  ggplot(aes(x=term, y=over_represented_pvalue) ) +
      geom_segment( aes(x=term ,xend=term, y=0, yend=over_represented_pvalue), color="grey") +
      geom_point(size=3, color="#69b3a2") +
      coord_flip() +
      theme(
        panel.grid.minor.y = element_blank(),
        panel.grid.major.y = element_blank(),
        legend.position="none"
      ) +
  xlab("") +
  ylab("over_represented_pvalue") +
  ggtitle("Adult") + #add a main title
  theme(plot.title = element_text(face = 'bold',
                                  size = 12,
                                  hjust = 0)) +
  theme_bw() + #Set background color
  theme(panel.border = element_blank(), # Set border
                     panel.grid.major = element_blank(), #Set major gridlines
                     panel.grid.minor = element_blank(), #Set minor gridlines
                     axis.line = element_line(colour = "black"), #Set axes color
        plot.background=element_blank(),#Set the plot background
        legend.position="none")
```



## 6. Session Info

```{r ressionInfo}
sessionInfo()
```


