---
title: "Pairwise Differential Gene Expression Analysis"
author: "Expression_Quant"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    keep_md: yes
---



```{r setup, include=FALSE}
# The following setting is important, do not omit.
knitr::opts_chunk$set(echo = TRUE,
                      fig.height = 8,
                      fig.width = 12,
                      dev = c("png", "pdf"),
                      dpi = 1000)
options(scipen = 999) # Prevent scientific notation
rm(list = ls()) # Clear environment
```

*Note: Open scripts.Rproj first, then script. To easily use relative paths, click the down button next to knit and then click "Knit Directory --> Project Directory". This should make loading and saving files much easier.*



# Differential Gene Expression Analysis

Perform pairwise Differential Gene Expression (DGE) analysis using a read count table (usually from Salmon) and a sample metadata sheet (user made).
Perform all pairwise DGE analyses, or limit to pairs provided by the user.
This script will output values for all genes (i.e., no significance filtering is applied) so that the user can filter downstream as needed.

<span style="color:red">**Red text**</span> indicates regions that require the user to modify.
Regardless, the user should check over all code blocks to ensure that everything is running correctly.



## 1. Load packages

Load packages.
```{r load_packages, message=FALSE, warning=FALSE}
library(tidyverse)
library(DESeq2)
library(genefilter)
```



## 2. Set variables

<span style="color:red">**Change file names and conditions where appropriate.**</span>
```{r variables}
# Input unfiltered data
treatmentinfo.file <- "samples.tsv"
gcount.file <- "salmon.numreads.tsv"
pairs2run.file <- "samples.pairs2run.tsv"
# Output DEG results
DGE_results.file <- "salmon.numreads.pairwise_DGE_results.tsv"

# Treatment condition column
condition <- "groupName"

# pOverA gene filtering thresholds
#  - We will only process genes with over these cutoffs.
#  - These cutoffs are only applied after selection for just the samples that we want to test (i.e., not the whole dataset)
pOverA.cutoff.P <- 0.4 # >40% of samples
pOverA.cutoff.A <- 10  # >10 read counts
```



## 3. Load, clean, and pre-processing datasets

Load the input file containing the treatment information.
```{r load_treatmentinfo}
treatmentinfo <- read.csv(treatmentinfo.file, header = TRUE, sep = "\t", fileEncoding="UTF-8-BOM") # Read in file
head(treatmentinfo)

# Check we have the right column names
headers <- c("sample_id", condition)
if( all(headers %in% colnames(treatmentinfo)) ){
  cat(paste(treatmentinfo.file, "has the required columns:", paste(headers, collapse=', ')))
} else {
  stop(paste(treatmentinfo.file, "is missing required columns:", paste(headers, collapse=', ')))
}
```


Load the input file containing the gene count matrix.
```{r load_gcount}
gcount <- as.data.frame(read.csv(gcount.file, header = TRUE, sep = "\t", fileEncoding="UTF-8-BOM")) # Read in file

# Check we have the right column names
headers <- c("Name", treatmentinfo$sample_id)
if( all(headers %in% colnames(gcount)) ){
  cat(paste(gcount.file, "has the required columns:", paste(headers, collapse=', ')))
} else {
  stop(paste(gcount.file, "is missing required columns:", paste(headers, collapse=', ')))
}

# Cleanup gcount data
gcount <- gcount %>%
  column_to_rownames("Name") %>% # Makes "Name" column rownames
  replace(is.na(.), 0) %>% # Replace missing values with zero
  round() %>% # Round
  select(treatmentinfo$sample_id) # Select just columns in filtered treatmentinfo file

# View dataset attributes
d <- dim(gcount); print(paste("rows:",d[[1]]," columns:",d[[2]], sep=''))
head(gcount)[,1:3]
```


Load the input file containing the pairs to run, if given.
```{r load_pairs2run}
## Function to handle detailed assertions
assert <- function(x, y, m){
  mm <- paste("'", paste(x[!x %in% y], collapse=','), "' ", m, ": ", paste(y, collapse=','), sep='')
  if(! all(x %in% y)){
    stop(mm)
  }
}

## Load treatment to compare - If given, load list of treatments to compare, else create all-vs-all list.
if(! is.null(pairs2run.file)) { # User has given a samples_combinations_file
  pairs2run <- as.data.frame(read.csv(pairs2run.file, header = TRUE, sep = "\t", fileEncoding="UTF-8-BOM", stringsAsFactors = FALSE)) # Read in file
  # Check that the user provided treatments are in the samples_info_file
  assert(pairs2run$Condition_1, treatmentinfo[, condition], "Treatment(s) in Condition_1 column of pairs2run.file are not in treatmentinfo.file")
  assert(pairs2run$Condition_2, treatmentinfo[, condition], "Treatment(s) in Condition_2 column of pairs2run.file are not in treatmentinfo.file")
} else { # Nothing given - Compute All-vs-all comparison
  tmp.Condition_1 <- c()
  tmp.Condition_2 <- c()
  treatments <- unique(treatmentinfo[, condition])
  for (i in 1:(length(treatments)-1)) {
    for (j in (i+1):length(treatments)) {
      tmp.Condition_1 <- c(tmp.Condition_1, treatments[i])
      tmp.Condition_2 <- c(tmp.Condition_2, treatments[j])
    }
  }
  pairs2run <- data.frame(Condition_1=tmp.Condition_1, Condition_2=tmp.Condition_2)
}


## Check we have the right column names
headers <- c("Condition_1", "Condition_2")
if( all(headers %in% colnames(pairs2run)) ){
  cat(paste(pairs2run.file, "has the required columns:", paste(headers, collapse=', ')))
} else {
  stop(paste(pairs2run.file, "is missing required columns:", paste(headers, collapse=', ')))
}

print("Comparisons to run:")
print(pairs2run)
```



## 4. Run Diff Expr analysis

Run all pairwise comparisons through DESeq2.
```{r run_DESeq2_pairs}
write(paste(c("Name", "baseMean", "log2FC", "lfcSE", "stat", "pvalue", "padj", "Condition_1", "Condition_2"), collapse='\t'), file=DGE_results.file, append=FALSE)
data.rownames <- rownames(gcount)
for (i in 1:nrow(pairs2run)){
  
  ## Get treatments to compare
  treatments <- pairs2run[i,]
  Condition_1 <- treatments$Condition_1
  Condition_2 <- treatments$Condition_2
  
  ## Print info of which treatments are being compared
  print(paste("Comparing treatments ", Condition_1, " **vs** ", Condition_2, sep=''))
  
  ## Extract samples for each treatment
  T1.samples=treatmentinfo[which(treatmentinfo[, condition]==Condition_1),]
  T2.samples=treatmentinfo[which(treatmentinfo[, condition]==Condition_2),]
  
  ## Extract accumulation data for selected samples
  T1.data <- gcount[,T1.samples$sample_id]
  T2.data <- gcount[,T2.samples$sample_id]
  
  ## Print info of which samples are being compared
  print(paste("Comparing samples ", paste(colnames(T1.data), collapse=','), " **vs** ", paste(colnames(T2.data), collapse=','), sep=''))
  
  ## Create filter for the counts data
  filt <- filterfun(pOverA(pOverA.cutoff.P, pOverA.cutoff.A))
  gfilt <- genefilter(T1.data, filt)
  T1.data.tokeep <- T1.data[gfilt,]
  
  filt <- filterfun(pOverA(pOverA.cutoff.P, pOverA.cutoff.A))
  gfilt <- genefilter(T2.data, filt)
  T2.data.tokeep <- T2.data[gfilt,]
  
  tokeep <- union(rownames(T1.data.tokeep), rownames(T2.data.tokeep))
  T1.data <- T1.data[tokeep,]
  T2.data <- T2.data[tokeep,]
  data <- cbind(T1.data, T2.data)
  
  ## Gene count data filtered in PoverA, P percent of the samples have counts over A
  print(paste("No. genes - before:",nrow(gcount)," after:",nrow(data), " (", round( (nrow(data)/nrow(gcount))*100, 2), "%)", sep=''))
  
  ## Create a sample sample design object for each column - “control” (“c”) or “treatment” (“t”)
  conditions <- c(rep("c", length(colnames(T1.data))), rep("t", length(colnames(T2.data))))
  coldata <- data.frame(row.names=c(colnames(T1.data), colnames(T2.data)), conditions, stringsAsFactors=TRUE)
  
  ## Create "dds" object for DESeq2 - contains count data and condition information
  dds <- DESeqDataSetFromMatrix(countData=data, colData=coldata, design=~conditions)
  
  ## Run DESeq2 - returns normalized expression level in logarithmic base 2 scale
  results <- results(DESeq(dds))
  
  ## Ensure results columns are in the correct order before we append them to output_file
  results <- results[c("baseMean", "log2FoldChange", "lfcSE", "stat", "pvalue", "padj")]
  
  ## Make sure every gene in the input matrix has a line in the results file.
  results <- as.data.frame(results)
  results$Name <- rownames(results)
  results <- merge(data.frame(Name=data.rownames), results, by="Name", all.x=TRUE, all.y=FALSE)
  
  ## Add treatments
  results$Condition_1 <- Condition_1
  results$Condition_2 <- Condition_2
  
  ## Append results to output_file
  write.table(results, file=DGE_results.file, append=TRUE, quote=FALSE, col.names=FALSE, row.names=FALSE ,sep='\t')
  
  ## Print any hidden warnings
  warnings()
  cat("\n\n")
}
warnings()
```



## 5. Session Info

```{r ressionInfo}
sessionInfo()
```


