---
title: "Detect Rhythmic Signals"
author: "Expression_Quant"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    keep_md: yes
---



```{r setup, include=FALSE}
# The following setting is important, do not omit.
knitr::opts_chunk$set(echo = TRUE,
                      fig.height = 8,
                      fig.width = 12,
                      dev = c("png", "pdf"),
                      dpi = 1000)
options(scipen = 999) # Prevent scientific notation
rm(list = ls()) # Clear environment
```

*Note: Open scripts.Rproj first, then script. To easily use relative paths, click the down button next to knit and then click "Knit Directory --> Project Directory". This should make loading and saving files much easier.*



# Detect Rhythmic Signals

Detect rhythmic signals using the `MetaCycle` package.

<span style="color:red">**Red text**</span> indicates regions that require the user to modify.
Regardless, the user should check over all code blocks to ensure that everything is running correctly.



## 1. Load packages

Load packages.
```{r load_packages, message=FALSE, warning=FALSE}
library(tidyverse)
library(MetaCycle)
```



## 2. Set variables

<span style="color:red">**Change file names and conditions where appropriate.**</span>
```{r variables}
# Input unfiltered data
treatmentinfo.file <- "samples.tsv"
gcount.file <- "salmon.tpm.tsv"
# Output MetaCycle results
rhythmic_results.file <- "salmon.numreads.rhythmic_gene_results.tsv"
rhythmic_results.sig.file <- "salmon.numreads.rhythmic_gene_results_significant.tsv"

# MetaCycle - expected period (24 = diurnal)
expected.period.length <- 24
expected.period.dev <- 4 # Allowed deviation from `expected.period.length` 20 -> 28

# Treatment condition column and treatments to include in analysis
condition <- "Group"
treatments.to.compare <- c("CTRL")

# pOverA gene filtering thresholds
#  - We will only process genes with over these cutoffs.
#  - These cutoffs are only applied after selection for just the samples that we want to test (i.e., not the whole dataset)
pOverA.cutoff.P <- 0.8 # >80% of samples
pOverA.cutoff.A <- 10  # >10 read counts

# Adjusted p-value cutoff for rhythmic genes
rhythmic.pvalue.cutoff <- 0.05
```



## 3. Load, clean, and pre-processing datasets

Load the input file containing the treatment information.
```{r load_treatmentinfo}
treatmentinfo <- read.csv(treatmentinfo.file, header = TRUE, sep = "\t", fileEncoding="UTF-8-BOM") # Read in file
head(treatmentinfo)

# Check we have the right column names
headers <- c("sample_id", condition)
if( all(headers %in% colnames(treatmentinfo)) ){
  cat(paste(treatmentinfo.file, "has the required columns:", paste(headers, collapse=', ')))
} else {
  stop(paste(treatmentinfo.file, "is missing required columns:", paste(headers, collapse=', ')))
}
```


Filter `treatmentinfo` to just the samples that we want to analyze.
<span style="color:red">**Adjust filtering as required.**</span>
```{r filter_treatmentinfo}
treatmentinfo <- treatmentinfo %>%
  filter(!!rlang::sym(condition) %in% treatments.to.compare) %>%
  mutate({{condition}} := factor(!!rlang::sym(condition), levels = treatments.to.compare))
head(treatmentinfo)
```


Load the input file containing the gene count matrix.
```{r load_gcount}
gcount <- as.data.frame(read.csv(gcount.file, header = TRUE, sep = "\t", fileEncoding="UTF-8-BOM")) # Read in file

# Check we have the right column names
headers <- c("Name", treatmentinfo$sample_id)
if( all(headers %in% colnames(gcount)) ){
  cat(paste(gcount.file, "has the required columns:", paste(headers, collapse=', ')))
} else {
  stop(paste(gcount.file, "is missing required columns:", paste(headers, collapse=', ')))
}

# Cleanup gcount data
gcount <- gcount %>%
  column_to_rownames("Name") %>% # Makes "Name" column rownames
  replace(is.na(.), 0) %>% # Replace missing values with zero
  round() %>% # Round
  select(treatmentinfo$sample_id) # Select just columns in filtered treatmentinfo file

# View dataset attributes
d <- dim(gcount); print(paste("rows:",d[[1]]," columns:",d[[2]], sep=''))
head(gcount)[,1:3]
```



## 4. Run Rhythmic Signal Detection

Run `meta2d` to select best algorithm to use for analysis.
```{r run_MetaCycle_pairs}
# Get the times of each sample in the same order as the columns in `gcount`
times <- treatmentinfo$Time_hrs
names(times) <- treatmentinfo$sample_id
timev <- times[colnames(gcount)]
names(timev) <- NULL

# Make `gcount` row names a column - required for `meta2d`
testD <- gcount %>% rownames_to_column("gene_id")

# Run Metacycle
cyc <- meta2d(inDF=testD, timepoints=timev,
              infile="csv", filestyle="csv",
              minper=(expected.period.length - expected.period.dev),
              maxper=(expected.period.length + expected.period.dev),
              ARSdefaultPer=expected.period.length,
              outputFile=FALSE, outRawData=TRUE
             )

# Print significant count
cyc.sig.meta <- cyc$meta[cyc$meta$LS_BH.Q < rhythmic.pvalue.cutoff, ]
print(paste("Num significant genes: ", nrow(cyc.sig.meta), sep=''))

# Write all results to file
write.table(cyc$meta, rhythmic_results.file, sep='\t', row.names=F, quote=F)

# Write significant results to file
write.table(cyc.sig.meta, rhythmic_results.sig.file, sep='\t', row.names=F, quote=F)
```



## 5. Session Info

```{r ressionInfo}
sessionInfo()
```


