---
title: "Differential Gene Expression Plot Genes"
author: "Expression_Quant"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    keep_md: yes
---



```{r setup, include=FALSE}
# The following setting is important, do not omit.
knitr::opts_chunk$set(echo = TRUE,
                      fig.height = 8,
                      fig.width = 12,
                      dev = c("png", "pdf"),
                      dpi = 1000)
options(scipen = 999) # Prevent scientific notation
rm(list = ls()) # Clear environment
```

*Note: Open scripts.Rproj first, then script. To easily use relative paths, click the down button next to knit and then click "Knit Directory --> Project Directory". This should make loading and saving files much easier.*



# Pairwise Differential Gene Expression Analysis

Code to plot accumulation values and highlight points with significant differential accumulation.

<span style="color:red">**Red text**</span> indicates regions that require the user to modify.
Regardless, the user should check over all code blocks to ensure that everything is running correctly.



## 1. Load packages

Load packages.
```{r load_packages, message=FALSE, warning=FALSE}
library(tidyverse)
library(reshape2)
library(cowplot)
library(ggsignif)
library(ggpubr)
```



## 2. Set variables

<span style="color:red">**Change file names and conditions where appropriate.**</span>
```{r variables}
# Input unfiltered data
treatmentinfo.file <- "samples.tsv"
gcount.file <- "salmon.tpm.tsv"
DGE_results.file <- "salmon.numreads.pairwise_DGE_results.tsv" # OPTIONAL: Just put `NULL` if no DEG results.

# Treatment conditions and variables to consider
condition.to.groupby <- "Group"    # e.g., Amb     - Used for coloring
condition.x.axis <- "Time_hrs"     # e.g., 12      - Used for spacing of points along x-axis
condition.persample <- "groupName" # e.g., Amb_12  - Used for separating points and coloring along x-axis
treatments.to.compare <- c("CTRL", "Glu_D", "Glu_L", "Lac_D", "Lac_L") # In `condition.to.groupby`
treatment.colors <- c("CTRL"="#33a02c",
                      "Glu_D"="#fdb462", "Glu_L"="#1f78b4",
                      "Lac_D"="#fb8072", "Lac_L"="#6a3d9a") # In `condition.to.groupby`
```



## 3. Load, clean, and pre-processing datasets

Load the input file containing the treatment information.
```{r load_treatmentinfo}
treatmentinfo <- read.csv(treatmentinfo.file, header = TRUE, sep = "\t", fileEncoding="UTF-8-BOM") # Read in file
head(treatmentinfo)

# Check we have the right column names
headers <- c("sample_id", condition.to.groupby, condition.x.axis)
if( all(headers %in% colnames(treatmentinfo)) ){
  cat(paste(treatmentinfo.file, "has the required columns:", paste(headers, collapse=', ')))
} else {
  stop(paste(treatmentinfo.file, "is missing required columns:", paste(headers, collapse=', ')))
}
```


Filter `treatmentinfo` to just the samples that we want to analyze. Keep commented out if you want to use everything.
<span style="color:red">**Adjust filtering as required.**</span>
```{r filter_treatmentinfo}
#treatmentinfo <- treatmentinfo %>%
#  filter(!!rlang::sym(condition) %in% treatments.to.compare) %>%
#  mutate({{condition}} := factor(!!rlang::sym(condition), levels = treatments.to.compare))
head(treatmentinfo)
```


Load the input file containing the gene count matrix.
```{r load_gcount}
gcount <- as.data.frame(read.csv(gcount.file, header = TRUE, sep = "\t", fileEncoding="UTF-8-BOM")) # Read in file

# Check we have the right column names
headers <- c("Name", treatmentinfo$sample_id)
if( all(headers %in% colnames(gcount)) ){
  cat(paste(gcount.file, "has the required columns:", paste(headers, collapse=', ')))
} else {
  stop(paste(gcount.file, "is missing required columns:", paste(headers, collapse=', ')))
}

# Cleanup gcount data
gcount <- gcount %>%
  column_to_rownames("Name") %>% # Makes "Name" column rownames
  replace(is.na(.), 0) %>% # Replace missing values with zero
  round() %>% # Round
  select(treatmentinfo$sample_id) # Select just columns in filtered treatmentinfo file

# View dataset attributes
d <- dim(gcount); print(paste("rows:",d[[1]]," columns:",d[[2]], sep=''))
head(gcount)[,1:3]
```


Melt and merge `gcount` and `treatmentinfo` into a final cleaned data.frame for plotting.
<span style="color:red">**Adjust the selected `TimePoints` and `Condition` as required.**</span>
```{r filter_gcount}
# Melt accumulation data.frame and merge with sample info. Will produce a data.frame with the following columns:
# Sample:          Name of sample
# Name:            Name of gene in sample
# Accumulation:    Accumulation value of gene in sample
# TimePoints:      `treatmentinfo` time points to use for x-axis (defined by user)
# Condition:       `treatmentinfo` variable to group accumulation values by at each time point (defined by user)
# TimePointLabel:  Label to use above each time point in plot (string; make whatever you want)
headers <- c("Sample", "Name", "Accumulation", "Condition", "TimePoints", "TimePointLabel")

gcount.merged <- gcount %>%
  as.matrix() %>%
  melt() %>% 
  rename(Name = Var1) %>%
  rename(Sample = Var2) %>%
  rename(Accumulation = value) %>%
  merge(treatmentinfo,
        by.x = "Sample",
        by.y = "sample_id") %>%
  mutate(TimePoints = TimePoint) %>%                                                  # USER ADJUST
  mutate(Condition = Group) %>%                                                       # USER ADJUST
  mutate(TimePointLabel = paste("TP", TimePoint, " (", Time_hrs, "hrs)", sep='')) %>% # USER ADJUST
  select(all_of(headers))

# Check we have the right column names
if( all(headers %in% colnames(gcount.merged)) ){
  cat(paste("gcount.merged has the required columns:", paste(headers, collapse=', ')))
} else {
  stop(paste("gcount.merged is missing required columns:", paste(headers, collapse=', ')))
}
```


Load the input file containing the DGE results to add to the plot (if given).
```{r load_DGE_results}
if(! is.null(DGE_results.file)) {
  diffExpr.data <- read.csv(DGE_results.file, header = TRUE, sep = "\t", fileEncoding="UTF-8-BOM") # Read in file
  
  # Check we have the right column names
  headers <- c("Name", "log2FC", "pvalue", "padj", "Condition_1", "Condition_2")
  if( all(headers %in% colnames(diffExpr.data)) ){
    cat(paste(DGE_results.file, "has the required columns:", paste(headers, collapse=', ')))
  } else {
    stop(paste(DGE_results.file, "is missing required columns:", paste(headers, collapse=', ')))
  }
  
  # Add colors and timepoints to diffExpr results
  t <- unique(treatmentinfo[, c(condition.persample, "TimePoint")])
  col <- treatmentinfo[c(condition.persample, condition.to.groupby)] %>%
    mutate(color = treatment.colors[!!rlang::sym(condition.to.groupby)]) %>%
    merge(t, by=condition.persample) %>%
    unique()
  
  diffExpr.data <- diffExpr.data %>%
    merge(col, by.x="Condition_1", by.y=condition.persample) %>%
    rename(Condition_1.Group = Group) %>%
    rename(Condition_1.color = color) %>%
    rename(Condition_1.TimePoint = TimePoint) %>%
    merge(col, by.x="Condition_2", by.y=condition.persample) %>%
    rename(Condition_2.Group = Group) %>%
    rename(Condition_2.color = color) %>%
    rename(Condition_2.TimePoint = TimePoint)
  head(diffExpr.data)
  
} else {
  diffExpr.data <- NULL
}
```



## 4. Plot Genes

<span style="color:red">**Adjust the plotting function as required.**</span>
```{r plot_results}
plot_results <- function(gene.id){
##
## Set envs
##
#gene.id <- "Gsulp_5572-G6470.1" # IDs of genes to plot
cutoff.FC=0.5                   # log2 Fold Change value to use for Up and Down abundance changes [abs(FC) > cutoff.FC]
cutoff.pvalue=NULL              # *p*-value to use for filtering differential abundance results [padjust < cutoff.pvalue]
cutoff.padj=0.05                # Adjusted *p*-value to use for filtering differential abundance results [padjust < cutoff.padj]
plot.colors=treatment.colors    # Colors to use for plotting
font.family="sans"              # Font family to use in plot
plot.legend=TRUE                # Add legend to plot?
with.points=TRUE                # Add replicate points to plot?
out.prefix=NULL                 # Prefix for output data tables (default: no data tables are output)

jitter.width <- 0.12            # Amount to jitter points over each box
timepoint.label.size <- 2       # Timepoint label size to use in plot
y.max.expand <- 1.05            # Expand y-axis by this amount (1.0 = limit is max accumulation value)

gene.id.data   <- gcount.merged %>% filter(Name==gene.id)
diffExpr.data <- diffExpr.data
plot.title    <- paste("Expression Results: ", gene.id, sep='')
xlab          <- "TimePoints"
ylab          <- "TPM"

signif.offset.x <- 0.1  # signif width of line either side of center
signif.offset.y <- 0.03 # signif spacing between values along y-axis
hjust <- 0.5            # signif offset text below bars
vjust <- 0.5            # signif offset text from center
textsize <- 2           # signif text size



##
## Make sure we know which p-value (normal or adjusted) to filter by
##
if(is.null(cutoff.pvalue)){
  p.column <- "padj"
  p.cutoff <- cutoff.padj
} else {
  p.column <- "pvalue"
  p.cutoff <- cutoff.pvalue
}



##
## Check that we have data (i.e., gene is in set)
##
if(nrow(gene.id.data) == 0){
  stop(paste(gene.id, "is not in gene set!"))
}
# Write data to file if file prefix given
if(!is.null(out.prefix)){
  write.table(gene.id.data,
              file=paste(out.prefix,".ExprData.tsv",sep=''),
              sep="\t",
              quote=FALSE,
              row.names=FALSE,
              col.names=TRUE)
}

## Get y-axis max values to set plot limits
y.min <- 0
y.max <- max(gene.id.data$Accumulation) * y.max.expand
if(y.max == 0){
  y.max <- 1
}



##
## Get DiffExpr results for gene.id for plot
##

if(! is.null(diffExpr.data)){
  t.diffExpr <- diffExpr.data %>%
    filter(abs(log2FC) > cutoff.FC) %>%
    filter(!!dplyr::sym(p.column) < p.cutoff) %>%
    filter(Name==gene.id)
  
  group.opts <- unique(c(t.diffExpr$Condition_1.Group, t.diffExpr$Condition_2.Group))
  t <- 1:length(group.opts)
  names(t) <- group.opts
  group.opts <- t
  
  t <- t.diffExpr %>%
    filter(Condition_1.TimePoint==Condition_2.TimePoint)
  
  t.signif.within <- data.frame(x          =c(t$Condition_1.TimePoint - (signif.offset.x * 2.5)),
                                xend       =c(t$Condition_1.TimePoint + (signif.offset.x * 2.5)),
                                y          =c(y.max + ((y.max * signif.offset.y) * group.opts[t$Condition_2.Group])),
                                annotation =sprintf("FC=%.2f; %s=%.2g", t$log2FC, p.column, t[[p.column]]),
                                color      =c(t$Condition_2.color)
  )
  
  signif.within <- geom_signif(stat = "identity", inherit.aes = FALSE,
                data = t.signif.within,
                aes(x = x,
                    xend = xend,
                    y = y,
                    yend = y,
                    annotation = annotation,
                    textsize=textsize,
                    vjust=vjust),
                col=t.signif.within$color,
                angle = 0,
                hjust=hjust)
  
  
  
  
  t <- t.diffExpr %>%
    filter(Condition_1.TimePoint!=Condition_2.TimePoint) %>%
    rownames_to_column("rownum") %>%
    mutate(rownum = as.numeric(rownum))
  
  t.signif.across <- data.frame(x          =c(t$Condition_1.TimePoint - (signif.offset.x * 2.5)),
                                xend       =c(t$Condition_2.TimePoint + (signif.offset.x * 2.5)),
                                y          =c(y.max + ((y.max * signif.offset.y) * length(group.opts) + 1) + ((y.max * signif.offset.y) * t$rownum)),
                                annotation =sprintf("FC=%.2f; %s=%.2g", t$log2FC, p.column, t[[p.column]]),
                                color      =c(t$Condition_2.color)
  )
  
  signif.across <- geom_signif(stat = "identity", inherit.aes = FALSE,
                data = t.signif.across,
                aes(x = x,
                    xend = xend,
                    y = y,
                    yend = y,
                    annotation = annotation,
                    textsize=textsize,
                    vjust=vjust),
                col=t.signif.across$color,
                angle = 0,
                hjust=hjust)
  
  y.max <- max(t.signif.across$y, t.signif.within$y) * y.max.expand
}




##
## Plot samples
##
plot.panel.1 <- gene.id.data %>%
  ggboxplot(x="TimePoints", y="Accumulation", color="Condition") +
    xlab(xlab) +
    ylab(ylab) +
    theme_classic() +
    theme(plot.title   = element_text(family = font.family, size=8, face="bold"),
                   axis.text.x  = element_text(family = font.family, size=10),
                   axis.text.y  = element_text(family = font.family, size=10, margin = margin(t = 0, r = 0, b = 0, l = 10)),
                   axis.title.x = element_text(family = font.family, size=12, face="bold"),
                   axis.title.y = element_text(family = font.family, size=12, face="bold"),
                   legend.text  = element_text(family = font.family, size=4),
                   legend.position="bottom",
                   panel.grid.major.x = element_line(linetype = "dashed", colour = "black", linewidth = 0.1),
                   ) +
    scale_color_manual(values = plot.colors) +
    #scale_x_continuous() +
    scale_y_continuous(limits = c(y.min, y.max), expand = c(0, NA))


##
## Add vertical dashed lines at each time point + timepoint labels
##
time <- unique(gene.id.data[, c("TimePoints", "TimePointLabel")])
time <- time[order(time$TimePoints, decreasing = FALSE),]

plot.panel.1 <- plot.panel.1 +
  geom_vline(xintercept = time$TimePoints,
                      linetype = rep("dashed", length(time$TimePoints)))

for(i in 1:nrow(time)) {
  row <- time[i,]
  # do stuff with row
  plot.panel.1 <- plot.panel.1 +
    geom_text(x = row$TimePoints, y = y.max, label = row$TimePointLabel,
                       size  = timepoint.label.size,
                       color = "black",
                       angle = 0,
                       vjust = 1.5,
                       hjust = -0.1
                      )
}



##
## Plot points
##
if(with.points){
  plot.panel.1 <- plot.panel.1 + geom_point(aes(color = Condition), position = position_jitterdodge(jitter.width=jitter.width))
}



##
## Add DiffExpr results to plot
##
if(! is.null(diffExpr.data)){
  plot.panel.1 <- plot.panel.1 +
    signif.within +
    signif.across
}



##
## Plot for full legend
##
if(plot.legend) {
  plot.Legend <- gene.id.data %>%
    ggplot(aes(x=TimePoints, y=Accumulation, group=interaction(TimePoints, Condition), color=Condition)) +
    geom_boxplot() +
    scale_color_manual(values = plot.colors, name = "Treatment") +
    theme_minimal() +
    theme(legend.position="right") # Needs to match 'guide-box-right'
  Legend <- get_plot_component(plot.Legend, 'guide-box-right', return_all = TRUE)
} else {
  Legend <- grid.rect(gp=gpar(col="white"))
}



##
## Arrange plots
##
p <- ggarrange(plotlist=list(
  plot.panel.1 + 
    theme(legend.position = "none"),
  Legend
  ),
  ncol=2,
  nrow=1,
  widths = c(6, 0.75),
  heights = c(6)
)

## Add title
p <- annotate_figure(p, top = text_grob(plot.title, 
               color = "black", face = "bold", size = 14))

## Add white background
p <- ggdraw(p) +
  theme(plot.background = element_rect(fill="white", color = NA))

p
}
```



```{r gene_plots}
plot_results("Gsulp_5572-G6364.1")
```



## 6. Session Info

```{r ressionInfo}
sessionInfo()
```


